Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-10-18 Multilevel Missing Data/Multilevel Missing Data.Rmd
Add sub-sections and references to paper outline

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Gerko Vink
    affiliation: 'Utrecht University'
    # To add another line, use \AND at the end of the previous one as above
  - name: Thomas Debray
    # address: |
    #   | Department of Statistics and Mathematics,
    #   | Faculty of Biosciences,
    #   | Universitat Autònoma de Barcelona
    affiliation: |
      | University Medical Centre Utrecht
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. AKA missing on so many levels.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Outline door Thomas

-	Why are standard (ad hoc) methods not well suited?

-	What is heterogeniety? (i.e. variability within studies vs. variability between studies). Random effects for intercept term, predictor effects, and/or variance residual error

- Systematic vs sporadic missingness

-	MICE approach: one-stage imputation versus two-stage imputation methods

-	Imputation of study level versus patient-level covariates

-	Handling MNAR

- Case study: metamisc::impact()

@@

- JOMO in mice

@@

- [https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119333784.ch18](Chapter: Dealing with Missing Data in an IPD Meta-analysis)

- [https://www.gerkovink.com/miceVignettes/Multi_level/Multi_level_data.html](Vignette: gerkovink.com/miceVignettes/Multi_level)

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht
  - name: Valentijn
    affiliation: University Medical Center Utrecht \AND
  - name: Gerko Vink
    affiliation: Utrecht University
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
      # | Julius Center for Health Sciences and Primary Care, 
      # | University Medical Center Utrecht, Utrecht, The Netherlands
    #affiliation2:
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. AKA missing on so many levels.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
library(tidyverse)

```

## Multilevel data

- What is clustering/multilevel data? In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering). What do we mean by clustering? In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc. In other fields: e.g. official stats clustering at country-level, or social sciences clustering at school-level (related to the sampling design).

- What is heterogeniety (i.e. variability within studies vs. variability between studies)? ADD: Figure to show difference between patient-level datapoints vs cluster-level datapoints. And different data frame formats.

- What methods are required to analyze multilevel data?	Random effects for intercept term, predictor effects, and/or variance residual error. Add references, e.g. @hox17.

## Missing data

- Why/where does missingness occur? I.e., not only patient-level but also cluster-level.

- How can we categorize this? Systematic vs sporadic missingness, see @resc13. ADD: visualization of systematic vs sporadic missingness. Within systematic we have always missing (same value per cluster) and non-measured variables (may differ per patient). TODO: adjust md pattern to match text.

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(pat = rep("Monotone", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
  tibble(pat = rep("Non-monotone", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
  mutate(across(everything(), as.factor))
ggplot(patterns, aes(x = name, y = rownr, fill = value, type = pat)) +
  geom_tile(color = "black") +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~pat) +
  theme_void() +
  theme(legend.position = "none") 
```

-	Why are standard (ad hoc) missing data methods not well suited?

- What other methods are available? General overview of approaches, see @audi18 @grun18. E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.

-	Additional problem addressed in this paper: handling MNAR.

## Aim of this paper

- Provide practical guidelines with code snippets for imputation of incomplete multilevel data.

- Case study options: metamisc::impact (IPD), GREAT (IPD).

# Workflows

## Modeling choices

- ADD: Section with more in-depth info on different data types and suitable methods. Add some equations, but not too much -> grow in complexity.

- Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeneiality) --> explain predictor matrix. Just mention that we should think about random intercept/slope, but just refer to other papers. This paper is just a software tutorial. --> Typically, at least random intercepts, but often random slopes as well. Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there. Sometimes there's little reason to assume some variable is affected by heterogeniety. We won't focus on congeneiality, that's too complex for 2-3 papers.

## Conditional models

- How to define the imputation model(s) in mice?

- What do the different implementations look like?

## Pooling

- Pooling 'regular' parameters vs more 'exotic' parameters (SE of residual errors, or autocorreletion)

- ADD: export mids objects to other packages like lme4 or coxme(?)

@@

- JOMO in mice --> on the side for now

- Additional levels of clustering

@@

### 2021-10-18 Multilevel Missing Data/Multilevel-Missing-Data.tex
Add sub-sections and references to paper outline

\author{
Hanne Oberman\\Utrecht University \And Gerko Vink\\Utrecht
University \And Thomas Debray\\University Medical Centre Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{\pkg{mice}: Multilevel}

@@

\Address{
    Hanne Oberman\\
    Utrecht University\\
    First line\\
Second line\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
      }

@@

\hypertarget{outline-door-thomas}{%
\section{Outline door Thomas}\label{outline-door-thomas}}

\begin{itemize}
\item
  Why are standard (ad hoc) methods not well suited?
\item
  What is heterogeniety? (i.e.~variability within studies
  vs.~variability between studies). Random effects for intercept term,
  predictor effects, and/or variance residual error
\item
  Systematic vs sporadic missingness
\item
  MICE approach: one-stage imputation versus two-stage imputation
  methods
\item
  Imputation of study level versus patient-level covariates
\item
  Handling MNAR
\item
  Case study: metamisc::impact()
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  JOMO in mice
\end{itemize}

\hypertarget{references}{%
\section{References}\label{references}}

\begin{itemize}
\item
  \href{Chapter:\%20Dealing\%20with\%20Missing\%20Data\%20in\%20an\%20IPD\%20Meta-analysis}{https://onlinelibrary.wiley.com/doi/abs/10.1002/9781119333784.ch18}
\item
  \href{Vignette:\%20gerkovink.com/miceVignettes/Multi_level}{https://www.gerkovink.com/miceVignettes/Multi\_level/Multi\_level\_data.html}
\end{itemize}

>>>>

\author{
Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
Medical Center Utrecht \And Valentijn\\University Medical Center Utrecht
\AND Gerko Vink\\Utrecht University \And Thomas Debray\\University
Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn, Gerko
Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{\pkg{mice}: Multilevel}

@@

\Address{
    Hanne Oberman\\
    Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
          }

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{multilevel-data}{%
\subsection{Multilevel data}\label{multilevel-data}}

\begin{itemize}
\item
  What is clustering/multilevel data? In this paper, we discuss grouped
  observations, not longitudinal data (within-patient clustering). What
  do we mean by clustering? In the medical field: Clustering by studies
  (IPDMA), hospitals in registries, multi-center studies etc. In other
  fields: e.g.~official stats clustering at country-level, or social
  sciences clustering at school-level (related to the sampling design).
\item
  What is heterogeniety (i.e.~variability within studies vs.~variability
  between studies)? ADD: Figure to show difference between patient-level
  datapoints vs cluster-level datapoints. And different data frame
  formats.
\item
  What methods are required to analyze multilevel data? Random effects
  for intercept term, predictor effects, and/or variance residual error.
  Add references, e.g. \citet{hox17}.
\end{itemize}

\hypertarget{missing-data}{%
\subsection{Missing data}\label{missing-data}}

\begin{itemize}
\item
  Why/where does missingness occur? I.e., not only patient-level but
  also cluster-level.
\item
  How can we categorize this? Systematic vs sporadic missingness, see
  \citet{resc13}. ADD: visualization of systematic vs sporadic
  missingness. Within systematic we have always missing (same value per
  cluster) and non-measured variables (may differ per patient). TODO:
  adjust md pattern to match text.
\end{itemize}

\begin{CodeChunk}

\begin{center}\includegraphics{Multilevel-Missing-Data_files/figure-latex/patterns-1} \end{center}

\end{CodeChunk}

\begin{itemize}
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What other methods are available? General overview of approaches, see
  \citet{audi18} \citet{grun18}. E.g., imputation of study level versus
  patient-level covariates, and one-stage imputation versus two-stage
  imputation methods.
\item
  Additional problem addressed in this paper: handling MNAR.
\end{itemize}

\hypertarget{aim-of-this-paper}{%
\subsection{Aim of this paper}\label{aim-of-this-paper}}

\begin{itemize}
\item
  Provide practical guidelines with code snippets for imputation of
  incomplete multilevel data.
\item
  Case study options: metamisc::impact (IPD), GREAT (IPD).
\end{itemize}

\hypertarget{workflows}{%
\section{Workflows}\label{workflows}}

\hypertarget{modeling-choices}{%
\subsection{Modeling choices}\label{modeling-choices}}

\begin{itemize}
\item
  ADD: Section with more in-depth info on different data types and
  suitable methods. Add some equations, but not too much -\textgreater{}
  grow in complexity.
\item
  Ideally we impute with random everything and heteroscedastic errors:
  most generic method (no worry about congeneiality) --\textgreater{}
  explain predictor matrix. Just mention that we should think about
  random intercept/slope, but just refer to other papers. This paper is
  just a software tutorial. --\textgreater{} Typically, at least random
  intercepts, but often random slopes as well. Refer to other papers for
  background, we'll focus just on the software implementation of the
  situations mentioned there. Sometimes there's little reason to assume
  some variable is affected by heterogeniety. We won't focus on
  congeneiality, that's too complex for 2-3 papers.
\end{itemize}

\hypertarget{conditional-models}{%
\subsection{Conditional models}\label{conditional-models}}

\begin{itemize}
\item
  How to define the imputation model(s) in mice?
\item
  What do the different implementations look like?
\end{itemize}

\hypertarget{pooling}{%
\subsection{Pooling}\label{pooling}}

\begin{itemize}
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorreletion)
\item
  ADD: export mids objects to other packages like lme4 or coxme(?)
\end{itemize}

@@

\begin{itemize}
\item
  JOMO in mice --\textgreater{} on the side for now
\item
  Additional levels of clustering
\end{itemize}

\renewcommand\refname{References}
\bibliography{../References/multilevelmice.bib}
